---
title: "How Does a Bike-Share Navigate Speedy Success?"
author: "Pelumi"
date: '2022-04-20'
output:
  pdf_document: default
  html_document: default
---

## Loading Packages
### Loading the tiddyverse,readr,ggplot2,lubridate package
```{r}
library(tidyverse)
library(readr)
library(ggplot2)
library(lubridate)
```

## Importing the data 
```{r}
may_2021_df <- read.csv("202104-divvy-tripdata.csv")
june_2021_df <- read.csv("202105-divvy-tripdata.csv")
july_2021_df <- read.csv("202106-divvy-tripdata.csv")
august_2021_df <- read.csv("202107-divvy-tripdata.csv")
september_2021_df <- read.csv("202108-divvy-tripdata.csv")
october_2021_df <- read.csv("202109-divvy-tripdata.csv")
november_2021_df <- read.csv("202110-divvy-tripdata.csv")
december_2021_df <- read.csv("202111-divvy-tripdata.csv")
january_2022_df <- read.csv("202112-divvy-tripdata.csv")
february_2022_df <- read.csv("202201-divvy-tripdata.csv")
march_2022_df <- read.csv("202202-divvy-tripdata.csv")
april_2022_df <- read.csv("202203-divvy-tripdata.csv")
```
## To check for the columns in the dataframes
```{r}
colnames(may_2021_df)
colnames(june_2021_df)
colnames(july_2021_df)
colnames(august_2021_df)
colnames(september_2021_df)
colnames(october_2021_df)
colnames(november_2021_df)
colnames(december_2021_df)
colnames(january_2022_df)
colnames(february_2022_df)
colnames(march_2022_df)
colnames(april_2022_df)
```

## Checking dataframe for inconsistencies
```{r}
glimpse(may_2021_df)
glimpse(june_2021_df)
glimpse(july_2021_df)
glimpse(august_2021_df)
glimpse(september_2021_df)
glimpse(october_2021_df)
glimpse(november_2021_df)
glimpse(december_2021_df)
glimpse(january_2022_df)
glimpse(february_2022_df)
glimpse(march_2022_df)
glimpse(april_2022_df)
```

## Merging the rows to make anaysis easier 

### Looking at the output,the columns in the dataframe are consistent. The next step is to merge the dataset into a single one. I'll be using the bind_row function to merge the dataframes into a single one
```{r}
bike_share_data <- bind_rows(may_2021_df, june_2021_df, july_2021_df, august_2021_df, september_2021_df, october_2021_df, november_2021_df, december_2021_df,january_2022_df, february_2022_df, march_2022_df, april_2022_df)
```

### Exploring the new dataset

```{r}
View(bike_share_data)
dim(bike_share_data)
glimpse(bike_share_data)
colnames(bike_share_data)
```

```{r}
tibble(bike_share_data)
```


### There are 5,723,532 observations, 13 variables, and null values in the dataframe.
## Checking for distinct ride type and memeber type using the unique function
```{r}
unique(bike_share_data$rideable_type)
unique(bike_share_data$member_casual)
```


## Checking for duplicate rides
```{r}
duplicated(bike_share_data$ride_id)

```
###Output shows no duplicated value

## Check for negative time values
```{r}
bike_share_data %>%
  filter(ended_at < started_at)%>%
  count()

```
### There are 145 negative time values
### Removing the negative values

```{r}
bike_share_data <- bike_share_data%>% filter(ended_at > started_at)
glimpse(bike_share_data)

```

## Removing Null values
```{r}
bike_share_data <- bike_share_data %>%
  na.omit()
glimpse(bike_share_data)
```

##Renaming the Variables
### Renaming the riderable_type variable and the member_casual variable and merging the start long and lat, and the end long and lat

```{r}
bike_share_data <- bike_share_data %>%
  rename(bike_type = rideable_type, customer_type = member_casual)
```
```{r}
bike_share_data <- bike_share_data %>%
unite(start_lat_lng, c("start_lat", "start_lng" ), sep = " ")
```
```{r}
bike_share_data <- bike_share_data %>%
  unite(end_lat_lng, c("end_lat", "end_lng"), sep = " ")
```
```{r}
glimpse(bike_share_data)
tibble(bike_share_data)
```
## Change the column with datetime values from chr to datetime
### Didnt chnage it earlier, cause there was a need to merge dataframes, but now that the dataframes have been merged into a single one, theres need to change it, to make it possible to work it.


```{r}
bike_share_data <- bike_share_data %>%
  mutate(started_at = ymd_hms(started_at), ended_at = ymd_hms(ended_at))
```
```{r}
glimpse(bike_share_data)
```
```{r}
View(bike_share_data)
```
### Viewing the dataframe after converting it from chr to datetime, realised there are a lot of null values, so avoid descripancies in calculations, removing the null values
```{r}
bike_share_data <- bike_share_data %>%
  na.omit()
```
```{r}
glimpse(bike_share_data)
```

## Calculating the day of the week each ride started
### be extracting the day of the week each ride started from the started_at variable

```{r}
bike_share_data <- bike_share_data %>%
  mutate(day_of_week = wday(started_at, label = TRUE, abbr = TRUE))
```
### be extracting the month from the started_at variable as well
```{r}
bike_share_data <- bike_share_data %>%
  mutate(month = month(started_at, label = TRUE, abbr = TRUE))
```
### extracting the year too
```{r}
bike_share_data <- bike_share_data %>%
  mutate(year = year(started_at))
```
```{r}
tibble(bike_share_data)
```

## Creating the start_time and end_time variable from the started_at and ended_at variable
```{r}
bike_share_data <- bike_share_data %>%
  mutate(start_time = format(started_at, "%H:%M:%S")) %>%
  mutate(end_time = format(ended_at, "%H:%M:%S"))
```
### Converting the time variable into the datatype
```{r}
bike_share_data <- bike_share_data %>%
 mutate(start_time = hms(start_time)) %>%
  mutate(end_time = hms(end_time))
```
```{r}
glimpse(bike_share_data)
```
```{r}
tibble(bike_share_data)
```

## Creating a variable that calculates the length of each ride

```{r}
bike_share_data$ride_length <- difftime(bike_share_data$ended_at, bike_share_data$started_at, units = "mins")
```
```{r}
View(bike_share_data)
```
```{r}
head(bike_share_data)
```

## ANALYSIS
### Number of rides per month
```{r}
rides_per_month_df <- bike_share_data %>%
  select(customer_type, month) %>%
  group_by(month, customer_type)%>%
  count() %>%
  arrange()
```
```{r}
head(rides_per_month_df)
```
### Graph  showing the number of Rides for each month
```{r}
ggplot(data = rides_per_month_df,  aes(x = month, y = n,  fill = month))  + geom_col() + facet_wrap(~customer_type) + labs(title = "Number of Rides per Month", x = "Month", y = "Count")
```



### Total Number of Casual members and Memberships
```{r}
df_members <- bike_share_data %>%
  select(customer_type) %>%
  group_by(customer_type) %>%
  count() %>%
  arrange()
```
```{r}
head(df_members)
```

### Graph showing Casual and Member Rider
```{r}
ggplot(data = bike_share_data) + geom_bar(mapping = aes(x = customer_type, fill = customer_type)) + labs(title = "Customer Type", x = "Customer type", y = "Count")
```


### Total Number of bike type per month
```{r}
bike_type_month <- bike_share_data %>%
  select(bike_type, month) %>%
  group_by(month, bike_type)%>%
  count() %>%
  arrange()
```
```{r}
head(bike_type_month)
```
### Graph of Bike type per Month
```{r}
ggplot(data = bike_share_data) + geom_bar(mapping = aes(x = bike_type, fill = bike_type)) + facet_wrap(~customer_type) + labs(title = "Bike Type per Month", x = "Bike_Type", y = "Count")
```

## Total number of Riders per Weekday
```{r}
df_weekday <- bike_share_data %>%
  select(customer_type, day_of_week) %>%
group_by(day_of_week, customer_type) %>%
  count()%>%
  arrange()
head(df_weekday)
```
### Graph of Number of Riders per weekday
```{r}
ggplot(data = bike_share_data) + geom_bar(mapping = aes(x = day_of_week, fill = day_of_week)) + facet_wrap(~customer_type) + labs(title = "Riders per Weekday", x = "Weekday", y = "Count")
```

### Maximun ride length
```{r}
bike_share_data %>%
 select(ride_length) %>%
  summarise(max_ride_length = max(ride_length))
```
### Minimum ride length
```{r}
bike_share_data %>%
  select(ride_length) %>%
  summarise(min_ride_length = min(ride_length))
```
 
 
## Average ride length
```{r}
bike_share_data %>%
  select(customer_type,ride_length) %>%
  summarise(mean_ride_length = mean(ride_length))
```

## Create a dataframe for only Casual riders
```{r}
df_casual <- bike_share_data %>%
  filter(customer_type == "casual")
```
### Minimum ride_length for Casual riders
```{r}
df_casual %>%
  select(ride_length) %>%
  summarise(min_casual = min(ride_length))
```
### Maximum ride length for Casual riders
```{r}
df_casual %>%
  select(ride_length) %>%
  summarise(max_casual = max(ride_length))
```
### Average ride length for Casual member
```{r}
df_casual %>%
  select(ride_length) %>%
summarise(mean_ride_length = mean(ride_length))
```


## Create a dataframe for only member riders

```{r}
df_member_rider <- bike_share_data %>%
  filter(customer_type == "member")
```
### Max ride_length for Member riders
```{r}
df_member_rider %>%
  select(ride_length) %>%
summarise(min_ride_length = min(ride_length))
```
### Average ride_length for Member riders
```{r}
df_member_rider %>%
  select(ride_length) %>%
summarise(mean_ride_length = mean(ride_length))
```


```




